Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2022-02-25 Manuscript/Manuscript.tex
as

Palmer amaranth's plasticity allows it to respond successfully to
environmental changes. Palmer amaranth demonstrates a high degree of
plasticity to light, temperature, water availability, and human
management (Jha et al., 2010). Palmer amaranth has an extended
germination period of up to \_\_\_\_\_ (Ward et al., 2013). Germination
of Palmer amaranth was triggered by 18 C soil temperature at 5 cm depth
(Keeley et al., 1987), and optimal germination and biomass production
occurred at 35/30 C day and night temperatures (Guo and Al-Khatib,
2003). Palmer amaranth emergence can be influenced by tillage and the
use of preemergence herbicides (Chahal et al., 2021), and may result in
weed germination shifts within a population, as documented in
\emph{Bassia scoparia} (Sbatella and Wilson, 2010). In an experiment
where Palmer amaranth was subjected to continuous water stress, it
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019).

Palmer amaranth's current global range includes agronomic cropland in
Italy (Milani et al., 2021), soybean producing regions of Brazil and
Argentina (Larran et al., 2017; Küpper et al., 2017), and the Cotton
Belt (Garetson et al., 2019; Bagavathiannan and Norsworthy, 2016) and
southern Great Plains of the United States (Kumar et al. (2020), Crespo
et al. (2016){]}. However, its range appears to be expanding steadily,
and in some cases rapidly, in the U.S. For example, in states with large
Palmer amaranth infestations its range is gradually pushing north as it
displaces other weeds in agronomic row crops (R Werle, personal
observation), aided by wildlife (Farmer et al., 2017) and agricultural
machinery. But there are other examples where it has suddenly become a
problem weed hundreds of miles away from known infestations. New
infestations in Michigan, Indiana and North Dakota appear to have
resulted from Palmer amaranth contaminated livestock feed. New
infestations in Iowa and Minnesota were associated with planting
contaminated pollinator habitat seed (Yu et al., 2021). Palmer amaranth
infestations have not been detected in Canada, but Palmer amaranth seeds
were found in sweet potato slips imported into the country (Page et al.,
2021). Global warming will create new opportunities for Palmer amaranth
invasion. Although agronomic crops in warm environments like Australia
and Sub-Saharan Africa are currently at greatest risk for Palmer
amaranth invasion, warming temperatures will reduce barriers that may
have limited Palmer amaranths spread into cooler climates like Canada
and Northern Europe (Kistner and Hatfield, 2018; Briscoe Runquist et
al., 2019).

There are many areas in the U.S. Corn Belt (North Central States) where
Palmer amaranth is not yet established, and its potential adaptability
is untested. Nonetheless, the rapid expansion of Palmer amaranth across
the Northern U.S. is concerning and warrants investigations on its
adaptability. We do know that Palmer amaranth caused yield loss in
Illinois soybean fields (Davis et al., 2015), but an Iowa study showed
that Palmer amaranth was not as well adapted as waterhemp
(\emph{Amaranthus tuberculatus}) to conditions in that geography (Baker,
2021). We also know that Palmer amaranth plants that establish shortly
after row-crop planting have a much greater impact on crop yield than
plants that emerge after the crop has produced several leaves (MacRae et
al., 2013). Understanding Palmer amaranth morphology and development
under different agroecosystems and across a wide geography can enhance
our knowledge of its adaptability, and may also aid in designing
effective tactics to limit is range expansion and minimize its negative
effects on row crops. The objective of this study was to investigate the
flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing in corn, soybean and bareground environments
across five locations in the North Central United States.

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes }
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control & glyphosate & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 &  &  & \\

 & Seeding rate & 88956 & 86487 &  &  & \\

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

 & Stage at 1 cohort & V2-3 &  &  &  & \\

 & Stage at 2 cohort & V6-7 &  &  &  & \\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 &  &  & \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha-1 &  &  &  & \\

@@

 & Variety & DSR-1950 & AG21X8 &  &  & \\

 & Seeding rate & 296400 & 321237 &  &  & \\

 & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

 & Stage at 1 cohort & V1-2 &  &  &  & \\

 & Stage at 2 cohort & V5-6 &  &  &  & \\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 &  &  & \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam &  &  & \\

 & Ratio (sand-clay-silt) & 10-64-26 & 17-58-25 &  &  & \\

 & pH & 6.6 & 6.5 &  &  & \\

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} S-metolachor, 1324 g ai ha; \textsuperscript{c} S-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha; \textsuperscript{d} saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha}\\
\end{tabular}}
\end{table}

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (Team, 2021).

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of \textless{} 0.00. The
average marginal effects showed that Palmer amaranth growing in corn
resulted in 14.8\% fewer females plants (Table 2). Moreover, increasing
a cumulative flowering unit doy increases the probability of having a
female plant by 0.4\% (Table 2 and Figure5A). A similar trend was
observed for weight (Figure 5B) and height (Figure 5C), where the
probability of being female increased 0.1\% and 0.2\% for each unit
increase of weight (g) and height (cm), respectively (Table 2).

\textbackslash begin\{table\}{[}!h{]}

\textbackslash caption\{\label{tab:unnamed-chunk-3}Average marginal
effects of Palmer amaranth gender logistic model. Factor pararemter
values (crop\_corn and 'crop\_bareground) are calculated relative to
soybean.\} \centering \fontsize{10}{12}\selectfont

\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

weight & 0.001 & 0.000 & 0.000 & 0.001 & 2.179 & 0.029\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} Average Marginal Effects. \textsuperscript{b} Standard Error.}\\
\end{tabular}

\textbackslash end\{table\}

@@

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). In our study, plants
growing from the first cohort accumulated 42\% more biomass when
compared to the second cohort. Therefore, Palmer amaranth plants growing
in the second cohort is likely to produce less seeds regardless the
crop. Our observation is consistent with the findings that the first
Palmer amaranth cohort produced 50\% more seeds per plant than Palmer
amaranth plants established six weeks later in bareground (Webster and
Grey, 2015). Still, seed production at the second cohort will replenish
the soil seedbank. Seed production and deposition in the seedbank is
also a key factor for species perpetuation (Menges, 1987). Palmer
amaranth can produce a hundred thousands seeds per plant (Schwartz et
al., 2016; Keeley et al., 1987), which can stay viable in the soil
seedbank for at least 36 months (Sosnoskie et al., 2013). Therefore,
preventing Palmer amaranth seed production or/and seed migration to
non-native habitats is an essential strategy to minimize the species
impact in agroecosystems (Davis et al., 2015).

Floral initiation depends on the complex interaction between a plant's
genetic makeup and environmental conditions (Lang, 1965). We observed a
significant disruption in flowering caused by surrounding vegetation and
cohort timing. Although all three treatments began flowering at the same
time in the first cohort (about doy 180), it took much less time for the
bareground treatment to reach 90\% flowering (34 days versus 40 for
soybean and 71 for corn). The lack of competition (and stress) in the
bareground resulted in plants that flowered earlier, and were they not
harvested, would have had longer reproductive periods, thereby producing
more seed. Similarly, in the second cohort, the window for all plants to
initiate flowering was much shorter for the bareground (28 days)
compared to Palmer amaranth in soybean (50 days) or corn (44 days).
Significantly, plants in the bareground treatment flowered one week
earlier than soybean, and almost two weeks earlier than corn, again
allowing for greater seed production per plant had they not been
harvested at initial flowering. In some cases, early flowering may
confer an evolutionary advantage, provided the plant has an
indeterminate habit and flowering does not restrict plant growth. In an
Iowa study, Palmer amaranth initiated flowering two weeks prior to
waterhemp (Baker, 2021), and this characteristic may partially explain
why Palmer amaranth can displace waterhemp where both occur in the same
agronomic field. In other cases, however, early flower initiation may be
a response to stressful conditions as a plant attempts to reproduce
before running out of resources. In a dioecious species like Palmer
amaranth, exerting stress on plants to manipulate flowering may be
beneficial in limiting seed production (McFarlane et al., 2018;
Schliekelman et al., 2005). For example, when growing under water
stress, there was a seven days flowering mismatch between male and
female plants (Mesgaran et al., 2021), which can minimize plant
outcrossing, reduce seed production and the exchange of resistant
alleles (Jhala et al., 2021). In our study we observed

An ecological approach to reduce Palmer amaranth seed production is
understanding plant biology, including flowering pattern. Our study
suggests that Palmer amaranth flowering was slightly influenced by crops
and cohort timings. Palmer amaranth growing in bareground and corn
resulted in the overall shortest and longest flowering window,
respectively. The shortest flowering window of second cohorts of Palmer
amaranth growing in bareground highlighted the impact of photoperiod on
flowering. When growing in soybean, Palmer amaranth flowering window was
similar to bareground for the first cohort but similar to corn at second
cohort timing. Plant flowering initiation is complex and depends on the
complex interaction between genetic makeup and the environmental
conditions (Lang, 1965). We hypothesize that when growing in high
competition (e.g., second cohort), Palmer amaranth plants tend to
initiate flowering early, as well as having an extended flowering
window. A study has shown that Palmer amaranth initiated flowering two
weeks prior to the native waterhemp in Iowa (Baker, 2021). Early flower
initiation could also be a plant strategy when growing in stressful
conditions. For example, when growing under water stress, early
flowering in Palmer amaranth resulted in a mismatch between female and
male plants by seven days (Mesgaran et al., 2021). A mismatch in Palmer
amaranth male and female flowering period can minimize plant
outcrossing, reducing plant seed production and/or exchange of resistant
alleles (Jhala et al., 2021). Sex dimorphism manipulation is considered
a potential ecological pest control strategy (McFarlane et al., 2018;
Schliekelman et al., 2005).

The mechanisms of gender-determination in plant species is intriguing
and has arouse the curiosity of many scientists, including Charles
Darwin (Darwin, 1888). In our study, the gender model performance was
decent (AUC 0.64) considering the biology of plant flowering. A 1:1 male
and female sex ratio is a general an evolutionarily stable strategy for
plant species perpetuation (Fisher, 1930). However, a slight deviation
from 1:1 sex ratio occurs in some dioecious species. For example, the
dioecious \emph{Halophila stipulacea} is a female-biased plant in its
native habitat, but the naturalized \emph{H. stipulacea} has a 1:1 ratio
(Nguyen et al., 2018). Naturalization of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant gender
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). We observed sexual dimorphism in
Palmer amaranth in response to surrounding vegetation and plant
morphological attributes. Our model estimated that late flowering,
heavier and taller Palmer amaranth plants slightly deviated from 1:1
ratio in favor of female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
plants invested more in leaf area and leaf dry weight under nutrient
deficiency (Korres et al., 2017). We observed more female plants in
soybean and bareground compared to corn. Palmer amaranth plants in the
corn were more stressed by interspecific competition as evidenced by
less biomass and a lower weight:height ratio. Sexual dimorphism is
documented in other dioecious species (Barrett and Hough, 2013). For
example, stronger female plant competition and greater male tolerance to
herbivory was reported in \emph{Spinacia oleracea} (Pérez-Llorca and
Sánchez Vilas, 2019). Research on candidate genes for sex determination
in \emph{Amaranthus} species are currently underway but is far from
complete (Montgomery et al., 2021, 2019). Further studies are also
needed to understand the ecological basis of Palmer amaranth flowering,
including plant behavior under climate change.

>>>>

Palmer amaranth's plasticity allows it to respond successfully to
environmental changes. Palmer amaranth demonstrates a high degree of
plasticity to light, temperature, water availability, and human
management (Jha et al., 2010). Palmer amaranth has an extended
germination period throughout the growing season (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). Palmer amaranth emergence can be influenced by tillage
and the use of preemergence herbicides (Chahal et al., 2021), and may
result in weed germination shifts within a population, as documented in
\emph{Bassia scoparia} (Sbatella and Wilson, 2010). In an experiment
where Palmer amaranth was subjected to continuous water stress, it
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019).

Palmer amaranth's current global range includes agronomic cropland in
Italy (Milani et al., 2021), soybean producing regions of Brazil and
Argentina (Larran et al., 2017; Küpper et al., 2017), and the Cotton
Belt (Garetson et al., 2019; Bagavathiannan and Norsworthy, 2016) and
southern Great Plains of the United States Crespo et al. (2016).
However, its range appears to be expanding steadily, and in some cases
rapidly, in the U.S. For example, in states with large Palmer amaranth
infestations its range is gradually pushing north as it displaces other
weeds in agronomic row crops (R Werle, personal observation), aided by
wildlife (Farmer et al., 2017) and agricultural machinery (Ward et al.,
2013). But there are other examples where it has suddenly become a
problem weed hundreds of miles away from known infestations. New
infestations in Michigan, Indiana and North Dakota appear to have
resulted from Palmer amaranth contaminated livestock feed. New
infestations in Iowa and Minnesota were associated with planting
contaminated pollinator habitat seed (Yu et al., 2021). Palmer amaranth
infestations have not been detected in Canada, but Palmer amaranth seeds
were found in sweet potato slips imported into the country (Page et al.,
2021). Global warming will create new opportunities for Palmer amaranth
invasion. Although agronomic crops in warm environments like Australia
and Sub-Saharan Africa are currently at greatest risk for Palmer
amaranth invasion, warming temperatures will reduce barriers that may
have limited Palmer amaranths spread into cooler climates like Canada
and Northern Europe (Kistner and Hatfield, 2018; Briscoe Runquist et
al., 2019).

There are many areas in the U.S. Corn Belt (North Central states) where
Palmer amaranth is not yet established, and its potential adaptability
is untested. Nonetheless, the rapid expansion of Palmer amaranth across
the Northern U.S. is concerning and warrants investigations on its
adaptability. We do know that Palmer amaranth caused yield loss in
Illinois soybean fields (Davis et al., 2015), but an Iowa study showed
that Palmer amaranth was not as well adapted as waterhemp
(\emph{Amaranthus tuberculatus}) to conditions in that geography (Baker,
2021). We also know that Palmer amaranth plants that establish shortly
after row-crop planting have a much greater impact on crop yield than
plants that emerge after the crop has produced several leaves (MacRae et
al., 2013). Understanding Palmer amaranth morphology and development
under different agroecosystems and across a wide geography can enhance
our knowledge of its adaptability, and may also aid in designing
effective tactics to limit is range expansion and minimize its negative
effects on row crops. The objective of this study was to investigate the
flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing in corn, soybean and bareground environments
across five locations in the North Central United States.

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes from Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control\textsuperscript{a} & \emph{S}-metolachor\textsuperscript{b} fb glyphosate\textsuperscript{c} /  & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{d} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor\textsuperscript{f} \\

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 & G03C84-5122-EZ1 & P1197AM & DKC64-34RIB (2018) / G10T63-3000GT (2019) \\

 & Seeding rate & 88956 & 86487 & 79000 & 86485 & 88,000 (2018) / 79,000 (2019)\\

 & Weed control & \emph{S}-metolachor fb glyphosate & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{e} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 & V6-7 & V6-7 & V2-3 & V4-5\\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 & V6-7 & V6-7 & V6-7 & V10-11\\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 & May 5, 2018 / May 14, 2019 & April 27, 2018 / May 3, 2019 & May 8, 2018 / April 26, 2019 \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha\textsuperscript{-1} &  &  & N (46-0-0) at 336 kg ha\textsuperscript{-1} & UAN-32, 202 kg N ha\textsuperscript{-1}\\

@@

 & Variety & DSR-1950 & AG21X8 & AG28X7 & 3017R2X (2018) / P31A22X (2019) & AG33X8 (2018) / CZ3601 (2019) \\

 & Seeding rate & 296400 & 321237 & 220000 & 370650 & 395000 (2018) / 370000 (2019) \\

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr + pyroxasulfone & glyphosate & glyphosate + \emph{S}-metolachor fb glyphosate & glyphosate + \emph{S}-metolachor \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 & V1-2 & V1-2 & V1-2 & V1-2 \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 & V5-6 & V5-6 & V5-6 & V7-8\\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 & May 5, 2018 / May 14, 2019 & May 7, 2018 / May 17, 2019 & May 8, 2018 / May 17, 2019 \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam & 2018: Mace silt loam / 2019: Kuma silt loam & Crete silty clay loam & \\

 & pH & 6.6 & 6.5 & 2018: 6.1 / 2019: 5.1 & 5.2 & \\

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 & 2018: 2 / 2019: 1.7 & 3.4 & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a}Hoe weeding; \textsuperscript{b}\emph{S}--metolachor, 1324 g ai ha\textsuperscript{-1}; \textsuperscript{c}glyphoste, 1262 g ae ha\textsuperscript{-1}; \textsuperscript{d}saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha\textsuperscript{-1}; \textsuperscript{e}\emph{S}-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha\textsuperscript{-1}; \textsuperscript{f}\emph{S}--metolachor, 1070 g ai ha\textsuperscript{-1}; \textsuperscript{g}glyphoste, 870 g ae ha\textsuperscript{-1}} \\
\end{tabular}}
\end{table}

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (R Core Team, 2021).

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of \textless{} 0.00. The
average marginal effects showed that Palmer amaranth growing in corn
resulted in 14.8\% fewer females plants (Table 2). Moreover, increasing
a cumulative flowering unit doy increases the probability of having a
female plant by 0.4\% (Table 2 and Figure5A). A similar trend was
observed for weight (Figure 5B) and height (Figure 5C), where the
probability of being female increased 0.1\% and 0.2\% for each unit
increase of weight and height, respectively (Table 2).

\caption{\label{tab:unnamed-chunk-3}Average marginal means of Palmer amaranth sex dimorphism logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME\textsuperscript{a} & SE\textsuperscript{b} & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

weight & 0.001 & 0.000 & 0.000 & 0.001 & 2.179 & 0.029\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} Average Marginal Effects. \textsuperscript{b} Standard Error.}\\
\end{tabular}
\end{table}

@@

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). In our study, plants
growing from the first cohort accumulated 42\% more biomass when
compared to the second cohort. Therefore, Palmer amaranth plants growing
in the second cohort is likely to produce less seeds regardless the
crop. Our observation is consistent with the findings that the first
Palmer amaranth cohort produced 50\% more seeds per plant than Palmer
amaranth plants established six weeks later in bareground (Webster and
Grey, 2015). Still, seed production at the second cohort will replenish
the soil seedbank. Seed production and deposition in the seedbank is
also a key factor for species perpetuation (Menges, 1987). Palmer
amaranth can produce a hundred thousands seeds per plant (Schwartz et
al., 2016; Keeley et al., 1987), which can stay viable in the soil
seedbank for at least 36 months (Sosnoskie et al., 2013). Therefore,
preventing Palmer amaranth seed production or/and seed migration to
non-native habitats is an essential strategy to minimize the species
impact in agroecosystems.

An ecological approach to reduce Palmer amaranth seed production is
understanding plant biology, including flowering pattern. Floral
initiation depends on the complex interaction between a plant's genetic
makeup and environmental conditions (Lang, 1965). We observed a
significant disruption in flowering caused by surrounding vegetation and
cohort timing. Although all three treatments began flowering at the same
time in the first cohort (about doy 180), it took much less time for the
bareground treatment to reach 90\% flowering (34 days versus 40 for
soybean and 71 for corn). The lack of competition (and stress) in the
bareground resulted in plants that flowered earlier, and were they not
harvested, would have had longer reproductive periods, thereby producing
more seed. Similarly, in the second cohort, the window for all plants to
initiate flowering was much shorter for the bareground (28 days)
compared to Palmer amaranth in soybean (50 days) or corn (44 days).
Significantly, plants in the bareground treatment flowered one week
earlier than soybean, and almost two weeks earlier than corn, again
allowing for greater seed production per plant had they not been
harvested at initial flowering. In some cases, early flowering may
confer an evolutionary advantage, provided the plant has an
indeterminate habit and flowering does not restrict plant growth. In an
Iowa study, Palmer amaranth initiated flowering two weeks prior to
waterhemp (Baker, 2021), and this characteristic may partially explain
why Palmer amaranth can displace waterhemp where both occur in the same
agronomic field. In other cases, however, early flower initiation may be
a response to stressful conditions as a plant attempts to reproduce
before running out of resources. In a dioecious species like Palmer
amaranth, exerting stress on plants to manipulate flowering may be
beneficial in limiting seed production (McFarlane et al., 2018;
Schliekelman et al., 2005). For example, when growing under water
stress, there was a seven days flowering mismatch between male and
female plants (Mesgaran et al., 2021), which can minimize plant
outcrossing, reduce seed production and the exchange of resistant
alleles (Jhala et al., 2021).

The mechanisms of gender-determination in plant species is intriguing
and has arouse the curiosity of many scientists, including Charles
Darwin (Darwin, 1888). In our study, the gender model performance was
decent (AUC 0.64) considering the biology of plant flowering. A 1:1 male
and female sex ratio is a general an evolutionarily stable strategy for
plant species perpetuation (Fisher, 1930). However, a slight deviation
from 1:1 sex ratio occurs in some dioecious species. For example, the
dioecious \emph{Halophila stipulacea} is a female-biased plant in its
native habitat, but naturalized \emph{H. stipulacea} has a 1:1 ratio
(Nguyen et al., 2018). Naturalization of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant gender
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). We observed sexual dimorphism in
Palmer amaranth in response to surrounding vegetation and plant
morphological attributes. Our model estimated that late flowering,
heavier and taller Palmer amaranth plants slightly deviated from 1:1
ratio in favor of female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
plants invested more in leaf area and leaf dry weight under nutrient
deficiency (Korres et al., 2017). We observed more female plants in
soybean and bareground compared to corn. Palmer amaranth plants in the
corn were more stressed by interspecific competition as evidenced by
less biomass and a lower weight:height ratio. Sexual dimorphism is
documented in other dioecious species (Barrett and Hough, 2013). For
example, stronger female plant competition and greater male tolerance to
herbivory was reported in \emph{Spinacia oleracea} (Pérez-Llorca and
Sánchez Vilas, 2019). Research on candidate genes for sex determination
in \emph{Amaranthus} species are currently underway but is far from
complete (Montgomery et al., 2021, 2019). Further studies are also
needed to understand the ecological basis of Palmer amaranth flowering,
including plant behavior under climate change.

### 2022-02-25 Manuscript/table.tex
as

 & Weed control\textsuperscript{a} & \emph{S}-metolachor\textsuperscript{b} fb glyphosate\textsuperscript{c} /  & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{d} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & \\

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 & G03C84-5122-EZ1 & P1197AM & \\

 & Seeding rate & 88956 & 86487 & 79000 & 86485 & \\

 & Weed control & \emph{S}-metolachor fb glyphosate & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{e} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 & V6-7 & V6-7 & V2-3 & \\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 & V6-7 & V6-7 & V6-7 & \\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 & May 5, 2018 / May 14, 2019 & April 27, 2018 / May 3, 2019 & \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha\textsuperscript{-1} &  &  & N (46-0-0) at 336 kg ha\textsuperscript{-1} & \\

@@

 & Variety & DSR-1950 & AG21X8 & AG28X7 & 2018: 3017R2X; 2019 P31A22X & \\

 & Seeding rate & 296400 & 321237 & 220000 & 370650 & \\

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr + pyroxasulfone & glyphosate & glyphosate + \emph{S}-metolachor fb glyphosate & \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 & V1-2 & V1-2 &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 & V5-6 & V5-6 &  & \\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 & May 5, 2018 / May 14, 2019 & May 7, 2018 / May 17, 2019 & \\

>>>>

 & Weed control\textsuperscript{a} & \emph{S}-metolachor\textsuperscript{b} fb glyphosate\textsuperscript{c} /  & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{d} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor\textsuperscript{f} \\

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 & G03C84-5122-EZ1 & P1197AM & DKC64-34RIB (2018) / G10T63-3000GT (2019) \\

 & Seeding rate & 88956 & 86487 & 79000 & 86485 & 88,000 (2018) / 79,000 (2019)\\

 & Weed control & \emph{S}-metolachor fb glyphosate & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{e} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & glyphosate + \emph{S}-metolachor \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 & V6-7 & V6-7 & V2-3 & V4-5\\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 & V6-7 & V6-7 & V6-7 & V10-11\\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 & May 5, 2018 / May 14, 2019 & April 27, 2018 / May 3, 2019 & May 8, 2018 / April 26, 2019 \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha\textsuperscript{-1} &  &  & N (46-0-0) at 336 kg ha\textsuperscript{-1} & UAN-32, 202 kg N ha\textsuperscript{-1}\\

@@

 & Variety & DSR-1950 & AG21X8 & AG28X7 & 3017R2X (2018) / P31A22X (2019) & AG33X8 (2018) / CZ3601 (2019) \\

 & Seeding rate & 296400 & 321237 & 220000 & 370650 & 395000 (2018) / 370000 (2019) \\

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr + pyroxasulfone & glyphosate & glyphosate + \emph{S}-metolachor fb glyphosate & glyphosate + \emph{S}-metolachor \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 & V1-2 & V1-2 & V1-2 & V1-2 \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 & V5-6 & V5-6 & V5-6 & V7-8\\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 & May 5, 2018 / May 14, 2019 & May 7, 2018 / May 17, 2019 & May 8, 2018 / May 17, 2019 \\
